---
title: "Reproducible Research Project Assignment 1"
author: "Claudio Marcello"
date: "Thursday, July 17, 2014"
output: html_document
---

# Loading and preprocessing the data

Show any code that is needed to

1. Load the data (i.e. read.csv() )
```{r}
# Load required libraries
library(plyr)
library(lattice)

# Set the C locale
lct <- Sys.getlocale("LC_TIME"); Sys.setlocale("LC_TIME", "C")
# 1. Load the data
act <- read.table("activity.csv", sep = ",", header = TRUE, stringsAsFactors=FALSE)
```

2. Process/transform the data (if necessary) into a format suitable for your analysis

```{r echo = TRUE}
# 2.1 Convert date field from chr to date
act$date <- as.Date(act$date, ,format = '%Y-%m-%d')
# 2.2 Convert interval (integer) to time (POSIXct);  
# use today's date, it is of no interest; only time is of interest.
act$time <- as.POSIXct(
        sprintf("%02d:%02d:00", act$interval %/% 100, act$interval %% 100), 
        format = "%H:%M:%S")
```


# What is the mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.

1. Make a histogram of the total number of steps taken each day
```{r DailyTotalNumberofSteps}
# summarize steps by date
day.sum <- ddply(act, .(date), summarise, 
                 steps = sum(steps, na.rm = TRUE))

histogram(~steps, data = day.sum, 
          main = "Histogram of Daily Total Number of Steps",
          xlab = "Daily Total Number of Steps")
```

2. Calculate and report the mean and median total number of steps taken per day
```{r}
day.avg <- data.frame(rbind(
        mean(day.sum$steps, na.rm = TRUE), 
        median(day.sum$steps, na.rm = TRUE)))
names(day.avg) <- ("steps")
row.names(day.avg) <- c("mean: ", "median: ")
day.avg
```


# What is the average daily activity pattern?

1.Make a time series plot (i.e.  type = "l" ) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r AverageDailyActivityPattern}
time.sum <- ddply(act, .(time), summarise, 
                  steps = round(mean(steps, na.rm = TRUE), 1))

# Ticks and labels for X axis
# From 0:00 to 24:00 in steps of 2 hours;
# Use time.sum$time and pad 24:00 at the end.
time.scale <- rbind(
        data.frame(
                at = time.sum$time[0:11 * 24 + 1], 
                labels = strftime(time.sum$time[0:11 * 24 + 1], format = "%H:%M")), 
        data.frame(
                at = as.POSIXlt("24:00:00", format = "%H:%M:%S"), 
                labels = "24:00"))

xyplot(steps ~ time, data = time.sum, type = "l", 
       scales = list(x = list(
               at = time.scale$at, 
               labels = time.scale$labels)), 
       main = "Average Number of Steps")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
time.max <- time.sum[which.max(time.sum$steps), 1:2]
time.max$time <- strftime(time.max$time, format = "%H:%M")
row.names(time.max) <- "maximum: "
time.max
```


# Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as  NA ). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with  NA s)
```{r}
ncases <- data.frame(c(
        as.data.frame(table(complete.cases(act)))$Freq,
        nrow(act)))
names(ncases) <- "Cases"
row.names(ncases) <- c("Incomplete: ", "Complete: ", "Total: ")
ncases
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r}
# Assuming the activity has similar pattern for each day of the week,
# 1. group records by day of the week and interval,
# 2. fill missing data with corresponding group average
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
act$steps2 <- act$steps
act$group <- paste(weekdays(act$date), act$interval)
act <- ddply(act, .(group), function(df) {
        df$steps2[is.na(df$steps2)] <- mean(df$steps2, na.rm=TRUE);
        return(df)
        })
```

4. Make a histogram of the total number of steps taken each day and 
```{r DailyTotalNumberofSteps(Modified)}
day.sum$steps2 <- ddply(act, .(date), summarise, 
                        steps2 = sum(steps2, na.rm = TRUE)
                        )$steps2
histogram(~steps2, data = day.sum, 
          main = "Histogram of Daily Total Number of Steps (Completed)", 
          xlab = "Daily Total Number of Steps")
```

Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
day.avg <- data.frame(rbind(
        c(mean(day.sum$steps, na.rm = TRUE), 
          mean(day.sum$steps2, na.rm = TRUE)), 
        c(median(day.sum$steps, na.rm = TRUE), 
          median(day.sum$steps2, na.rm = TRUE))))
names(day.avg) <- c("steps", "steps2")
row.names(day.avg) <- c("mean", "median")
day.avg$increment <- with(day.avg, sprintf("%1.1f%%", (steps2 - steps)/ steps * 100))
day.avg
```


# Are there differences in activity patterns between weekdays and weekends?

For this part the  weekdays()  function may be of some help here. Use the dataset with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
act$dayType <- ordered(
        weekdays(act$date) %in% c("Sunday", "Saturday"),
        labels = c("weekday", "weekend"))
```

2. Make a panel plot containing a time series plot (i.e.  type = "l" ) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data:

```{r AverageWeeklyActivityPattern}
week.sum <- ddply(act, .(dayType, time), summarise, 
                  steps2 = round(mean(steps2), 1))
xyplot(steps2 ~ time | dayType, data = week.sum, 
       type = "l", layout = c(1, 2), 
       scales = list(x = list(
               at = time.scale$at, 
               labels = time.scale$labels)), 
       main = "Average Daily Activity Pattern", 
       ylab = "steps")
```

# --- End ---